---
title: 'R Tips: Looping'
author: "PSCI 8357, Spring 2016"
date: "January 22, 2016"
output:
  pdf_document:
    highlight: tango
    template: ../../charter-template.tex
  html_document:
    highlight: tango
    keep_md: yes
---

```{r setup, include=FALSE}
options(width = 70, digits = 3)
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, fig.align = "center")
```

In this week's assignment---and, most likely, future ones---you will want to perform the same operation on many different values.  We typically do this in R (and any other programming language) with a *loop*.  This brief note will introduce you to the concept of loops and illustrate some especially convenient ways to write loops in R.


## `for` Loops

Use a loop when you have:

1. A list of values
2. An operation to perform on each of those values

For example, suppose you want to:

* Draw $N = 5$ values from a standard normal distribution
* Calculate their maximum and save it
* Repeat for $N = 10, 15, 20, 25$

So you have the list of values $N = 5, \ldots, 25$ and the operation "draw $X_1, \ldots, X_N$ from $N(0, 1)$ and calculate their mean".  One way you could do this is the copy-and-paste way:

```{r copy-paste}
max_5 <- max(rnorm(5))
max_10 <- max(rnorm(10))
max_15 <- max(rnorm(15))
max_20 <- max(rnorm(20))
max_25 <- max(rnorm(25))

c(max_5, max_10, max_15, max_20, max_25)
```

Don't do this.  First, it doesn't scale well.  Imagine how much time you'd waste if you had to copy and paste this a hundred, or a thousand, or 300 million times.  Second, it's error prone.  At some point you *will* mess up and enter the wrong thing.  Third, it's a nightmare if the operation changes---you'll have to go back and manually fix every line of code.

A better way would be to write a loop, like in the following example.

```{r loop-example}
## Define the sequence of values
n_seq <- seq(5, 25, by = 5)

## Make a vector to store the results
max_by_n <- rep(NA, length(n_seq))

for (i in 1:length(n_seq)) {
    ## Retrieve the corresponding value
    n <- n_seq[i]

    ## Perform the operation and store the results
    x <- rnorm(n)
    max_by_n[i] <- max(x)
}

max_by_n
```

When we write `for (i in vector) { operation }`, we are telling R to take each individual value of `vector`, call it `i`, and perform `operation` on it.  So here, we take each value of `i` between 1 and 5, find the corresponding value of $N$, draw $N$ observations from a standard normal distribution, and save their maximum.


# `foreach` Loops

The code above is a bit clunky.  Before we write our loop, we have to set up storage for the results.  Instead of looping directly over $N$, we loop over the number of different values of $N$ we'll be examining.  We can write loops more elegantly using the [**foreach**](https://cran.r-project.org/web/packages/foreach/index.html) package.

```{r foreach, message=FALSE}
library("foreach")
```

The basic syntax of a foreach loop is `foreach (i = vector) %do% { operation }`.  Notice that we've replaced `for` with `foreach`, replaced `in` with `=`, and added `%do%` between the loop setup and the operation.  But there is an importance difference between `for` and `foreach`.

* `for` just performs the operation you ask it to.  If you don't tell it to save the results of the operation, it'll perform it and then forget.
* `foreach` saves the result of each operation it performs---specifically, the value of the last line of the operation---and returns the results as a list.

So we can rewrite our earlier loop much more elegantly using `foreach`.  The exact values of the output will be different than before since we're drawing random numbers, but this code performs the same task:

```{r foreach-translate}
n_seq <- seq(5, 25, by = 5)

max_by_n <- foreach (n = n_seq) %do% {
    max(rnorm(n))
}

max_by_n
```

Notice that we don't have to explicitly set up a place to store the results, and we can loop over $N$ instead of $1, \ldots, 5$.  The drawback is that the results are stored as a list, and we'd like a vector.  We can accomplish that by using the `.combine` argument to `foreach`:

```{r foreach-combine}
n_seq <- seq(5, 25, by = 5)

max_by_n <- foreach (n = n_seq, .combine = "c") %do% {
    max(rnorm(n))
}

max_by_n
```

For operations that spit out a vector of numbers, you can stack them using `.combine = "rbind"` or place them in columns with `.combine = "cbind"`.  Both of these assume that each vector of output will be the same length---if not, then you'll have to take the output as a list.


# When to Avoid Looping

For many common operations, you don't have to write a loop---R does it for you.  For example, imagine you want to take the square root of each of a vector of numbers.  The `sqrt()` function automatically loops over them.

```{r sqrt}
y <- 1:10
sqrt(y)
```

We call functions like `sqrt()` *vectorized*, because they operate on each element of a vector.  Before using a `for` loop, always check for a vectorized function that does what you want.
